---
title: "GBA imputation part 2"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 7
theme: cosmo
---

Source: https://github.com/markziemann/impute2/

## Introduction

IDEA:

* Collect the correlation coefficients of genes in a set (called GS).

* Use the background distribution for comparison.

* For each gene not in the set, ask whether the correlation to genes in the set is similar to genes in the set or it looks different.

* Get a probability that the gene is part of the set or not.


## Libraries

```{r,packages}

suppressPackageStartupMessages({
  library("tictoc")
  library("parallel")
  library("vioplot")
  library("fgsea")

  library("reshape2")
  library("eulerr")
  library("HGNChelper")
  library("kableExtra")
  library("beeswarm")
  library("gplots")
  library("gridExtra")
  library("png")
  library("parallel")
  library("RhpcBLASctl")
  library("tictoc")
  library("vioplot")
})

# geometric mean https://stackoverflow.com/questions/2602583/geometric-mean-is-there-a-built-in
gm_mean = function(x, na.rm=TRUE){
  exp(sum(log(x[x > 0]), na.rm=na.rm) / length(x))
}

gmt_import <- function(gmtfile) {
    genesetLines <- strsplit(readLines(gmtfile), "\t")
    genesets <- lapply(genesetLines, utils::tail, -2)
    names(genesets) <- unlist(lapply(genesetLines, head, 1))
    attributes(genesets)$originfile <- gmtfile
    genesets <- genesets[lapply(genesets,length)>0]
    if( any(duplicated(names(genesets))) ) {
        warning("Duplicated gene sets names detected")
    }
    genesets
}

```

## Load gmt

Selecting three cell cycle GO BPs with different sizes (small, medium and large).


```{r,gmt1}

#go <- gmtPathways("c5.go.v2024.1.Hs.symbols.gmt")
go <- gmt_import("c5.go.v2024.1.Hs.symbols.gmt")

summary(unlist(lapply(go,length)))
table(unlist(lapply(go,length))>500)

gs <- go[which(unlist(lapply(go,length))<=500)]

#gs <- go[grep("CELL_CYCLE",names(go))][1:3]

head(lapply(gs,length))
summary(unlist(lapply(gs,length)))



```

## Get the gene symbol - ENSG ID relationships


```{r,ensg1}

ensg <- read.table("ensg.tsv",header=TRUE)

```

## Index the location of genes in the cordata

GX2 is an object with gene ID, file number and column number.
This will allow us to extract those columns and start analysing the correlation
coefficients.

```{r,cordata1}

cdatlist <- list.files(".",pattern="^cordata")
gx1 <- lapply(cdatlist, function(f) {
  h1 <- readLines(f,n=1)
  h1 <- unlist(strsplit(h1,'\t'))
  return(h1)
} )

names(gx1) <- 1:length(gx1)
gx2 <- stack(gx1)

gx3 <- lapply(gx1, function(f) { 1:length(f) } )

names(gx3) <- 1:length(gx1)

gx4 <- stack(gx3)

gx2$pos <- gx4$values

head(gx2)

```

## Index the row names

```{r,rownames1}

rw <- read.table("rownames.txt",header=FALSE)$V1
head(rw)

```

## Get the genes and pearson correlation coefficients

Need to select specific columns.

This seems to work pretty well for small gene sets by finding genes that have high correlation coefficients
to genes already in the set.

It looks at the relationship of an unknown gene to all the genes in the set.
This may work for smaller sets, but for larger pathways it might be unreasonable
to expect a gene to have high correlation with 100+ members of that set.

This might work better if we set a threshold such as selecting the coefficients of
unknown gene with the top 10 best genes already in the set.
Instead of a hard cut-off, a weighting could be better.
For example if there are 100 genes in the set, then



```{r,extract1}

#gs <- head(gs,100)

tic()
results <- lapply(1:length(gs), function(i) {
  mygs <- gs[i]
  gsname <- names(mygs)
  message(paste("No.",i,"of 10086:",gsname))
  gs_genes <- unname(unlist(mygs))
  gs_ensg <- ensg[match(gs_genes,ensg$hgnc_symbol),1]
  gx2$values <- gsub("[^[:alnum:] ]", "", gx2$values)
  gs_gx <- gx2[gx2$values %in% gs_ensg,]
  if ( nrow(gs_gx) < 5 ) { return(NULL) }
  message("extracting columns")
  dfl <- mclapply(unique(gs_gx$ind), function (f) {
    mypos <- gs_gx[gs_gx$ind == f,3]
    mypos2 <- paste(mypos,collapse=",")
    TMPFILE=paste("tmp.",letters[f],sep="")
    COMMAND <- paste("zcat", cdatlist[f], "| cut -f",mypos2,">",TMPFILE)
    system(COMMAND)
    cx <- read.table(TMPFILE,header=TRUE)
    return(cx)
  },mc.cores=16)
  message("Tidy output")
  cmx <- do.call(cbind,dfl)
  pcin <- cmx[which(rw %in% gs_ensg),]
  pcin <- as.numeric(unname(unlist(as.vector(pcin))))
  pcin <- pcin[which(pcin!=1)]
  pcout <- cmx[which(! rw %in% gs_ensg),]
  pcout <- as.numeric(unname(unlist(as.vector(pcout))))
  pcout <- pcout[which(pcout!=1)]
  mid_in <- mean(c(mean(pcin),median(pcin)))
  mid_out <- mean(c(mean(pcout),median(pcout)))
  message("now test each gene for membership") #need more ram for big gene sets
  gm <- unlist(mclapply(1:nrow(cmx), function(i) {
    gpc <- as.numeric(unname(unlist((cmx[i,,drop=TRUE]))))
    gpc <- gpc[which(gpc!=1)]
    mid <- mean(c(mean(gpc),median(gpc)))
    return(mid)
  }, mc.cores=24))
  message("tidy output")
  names(gm) <- rw
  candidates <- tail(sort(gm),(length(gs_ensg)*2))
  candidates <- candidates[which(candidates > mid_in)]
  candidate_symbols <- ensg[match(names(candidates),ensg$ensembl_gene_id),2]
  res <- data.frame(names(candidates), unname(candidates),candidate_symbols,mid_out,mid_in,gsname)
  colnames(res) <- c("ensID","cand_r","geneSymbol","out_r","in_r","setName")
  res <- res[,c("ensID","geneSymbol","setName","cand_r","in_r","out_r")]
  return(res)
})
toc()
head(results)
length(results)

```

## Investigate weighting

As some gene sets are large and it would be unreasonable to expect a gene to correlate with all genes in the set,
we can potentially investigate weighting, which would emphasise tight co-regulation with the top X partners.

Here are three approaches for comparison.

1. First, unweighted; all values count equally.

2. Linear, higher values are boosted at a linear degree.

3. Square parabola, the linear weighting factor is squared.

| Sorted value | Unweighted | Linear | square | Sq Parabola |
| --- | --- | --- | --- | --- |
| 1 | 1 | 1.9 | 3.61 | 2.71 |
| 2 | 1 | 1.7 | 2.89 | 2.17 |
| 3 | 1 | 1.5 | 2.25 | 1.69 |
| 4 | 1 | 1.3 | 1.69 | 1.27 |
| 5 | 1 | 1.1 | 1.21 | 0.91 |
| 6 | 1 | 0.9 | 0.81 | 0.61 |
| 7 | 1 | 0.7 | 0.49 | 0.19 |
| 8 | 1 | 0.5 | 0.25 | 0.07 |
| 9 | 1 | 0.3 | 0.09 | 0.07 |
| 10 | 1 | 0.1 | 0.01 | 0.01 |

## Session information

For reproducibility.

```{r,session}

sessionInfo()

```
