---
title: "DEE2 data prep"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 7
theme: cosmo
---

Source: https://github.com/markziemann/impute2/

## Introduction

```{r,packages}

suppressPackageStartupMessages({
  library("tictoc")
  library("parallel")
  library("vioplot")
  library("fgsea")

  library("reshape2")
  library("eulerr")
  library("HGNChelper")
  library("kableExtra")
  library("beeswarm")
  library("gplots")
  library("gridExtra")
  library("png")
  library("parallel")
  library("RhpcBLASctl")
  library("tictoc")
  library("vioplot")
})

```

Read in the QC data so that we can find the PASS datasets.

```{r,readqc}

download.file("https://dee2.io/mx/hsapiens_qc.tsv.bz2",destfile="hsapiens_qc.tsv.bz2")
q <- read.table("hsapiens_qc.tsv.bz2",sep="\t")
pass <- unique(subset(q,V3=="PASS"))
srr <- pass$V1
length(srr)
sname <- paste("/mnt/md0/dee2/data/hsapiens/",srr,"/",srr,".se.tsv.gz",sep="");
writeLines(sname,con="sname.txt")

```

Retabulate the individual STAR expression counts file into a transposed matrix.
Do RPM based normalisation.
We use this approach because reformatting the 3 column bz file requires >700 GB of memory,
which leaves none for downstream correlation analysis.

Takes 81m. File size: 16 GB, when gz compressed: 5.0 GB . Data rows: 113,332.

```{bash,retabulate}

MYSNAME=$(head -1 sname.txt)
zcat $MYSNAME | cut -f1 | paste -s -d '\t' | sed 's/DRR000177//' > colnames.tsv

retab() {
N=$1
SUM=$(zcat $N | cut -f2 | numsum)
SRR=$(echo $N | cut -d '/' -f7 )

zcat $N \
  | cut -f2 \
  | sed 1d \
  | awk -v S=$SUM '{print 1000000*$1/S}' \
  | paste -s -d '\t' \
  | sed "s/^/${SRR}\t/"

}
export -f retab

cat colnames.tsv > dat.tsv
parallel -j16 retab :::: sname.txt >> dat.tsv

```

Split into chunks. 1m26s.

```{bash,corr1}

cut -f1 dat.tsv > rownames.txt &
cut -f1-3645 dat.tsv > dat.tsv.a &
wait
cut -f3646-7289 dat.tsv | paste rownames.txt - > dat.tsv.b &
cut -f7290-10933 dat.tsv | paste rownames.txt - > dat.tsv.c &
cut -f10934-14577 dat.tsv | paste rownames.txt - > dat.tsv.d &
cut -f14578-18221 dat.tsv | paste rownames.txt - > dat.tsv.e &
cut -f18222-21865 dat.tsv | paste rownames.txt - > dat.tsv.f &
cut -f21866-25509 dat.tsv | paste rownames.txt - > dat.tsv.g &
cut -f25510-29153 dat.tsv | paste rownames.txt - > dat.tsv.h &
cut -f29154-32797 dat.tsv | paste rownames.txt - > dat.tsv.i &
cut -f32798-36441 dat.tsv | paste rownames.txt - > dat.tsv.j &
cut -f36442-40085 dat.tsv | paste rownames.txt - > dat.tsv.k &
cut -f40086-43729 dat.tsv | paste rownames.txt - > dat.tsv.l &
cut -f43730-47373 dat.tsv | paste rownames.txt - > dat.tsv.m &
cut -f47374-51017 dat.tsv | paste rownames.txt - > dat.tsv.n &
cut -f51018-54661 dat.tsv | paste rownames.txt - > dat.tsv.o &
cut -f54662-58305 dat.tsv | paste rownames.txt - > dat.tsv.p &
wait

```

Single file correlations.

Then multifile.

```{r,correlation1}

library("parallel")
flist <- list.files(".",pattern="dat.tsv.")
null <- mclapply(flist,function(f) {
  dat <- read.table(f,header=TRUE,row.names=1)
  outfile <- gsub("dat","cor",f)
  mycor <- cor(dat) #20 mins 10 GB RAM
  write.table(x=format(mycor,digits=5), file=outfile,sep="\t")
  return(0)
},mc.cores=16)

flist <- list.files(".",pattern="dat.tsv.")
fcomb <- expand.grid(flist,flist)
fcomb <- subset(fcomb,Var1!=Var2)
fcomb <- t(apply(fcomb,1,function(x) { x[order(x)] } ))
fcomb <- unique(fcomb)
length(fcomb)

null <- mclapply(1:nrow(fcomb), function(i) {
  aname <- as.character(fcomb[i,1])
  bname <- as.character(fcomb[i,2])
  altr <- sapply(strsplit(aname,"\\."),"[[",3)
  bltr <- sapply(strsplit(bname,"\\."),"[[",3)
  outname <- paste("cor.tsv.",altr,bltr,sep="")
  a <- read.table(aname,header=TRUE,row.names=1)
  b <- read.table(bname,header=TRUE,row.names=1)
  abcor <- cor(a,b) #XX miins 20 GB RAM
  write.table(x=format(abcor,digits=5), file=outname,sep="\t")
  return(0)
},mc.cores=20)

```

```{bash,realign_headers}

for COR in cor.tsv* ; do
  sed -i '1s/^/\t/' $COR
done

```

Reassemble the data.

```{r,cor}

myfiles <- list.files(".",pattern="cor.tsv")

ltrs <- letters[1:16]
null <- mclapply(ltrs, function(ltr) {
  argx <- paste("tsv.",ltr,sep="")
  afiles <- myfiles[grep(argx,myfiles)]
  brgx <- paste(ltr,"$",sep="")
  bfiles <- myfiles[grep(brgx,myfiles)]
  bfiles <- setdiff(bfiles,afiles)
  adata <- mclapply(afiles,function(x) { read.table(x) }, mc.cores=1)
  adata <- do.call(cbind, adata)
  bdata <- mclapply(bfiles,function(x) { read.table(x) }, mc.cores=1)
  bdata <- do.call(rbind, bdata)
  cordata <- rbind(t(adata),bdata)
  cordata <- cordata[order(rownames(cordata)),]
  cordata <- cordata[,order(colnames(cordata))]
  outfile <- paste("cordata.tsv.",ltr,sep="")
  write.table(x=cordata,file=outfile,sep="\t")
  return(0)
}, mc.cores=16)

```

Thresholding.

```{r,thresholding}

library("vioplot")
cordata <- read.table("cordata.tsv.a",header=TRUE)
rn <- row.names(cordata)
cordata <- apply(cordata,2,as.numeric)
rownames(cordata) <- rn
cordata[cordata == 1] <- NA
par(mar= c(10.1, 4.1, 4.1, 2.1) )
vioplot(cordata[,1:20],las=2,main="Correlation coefficients")
par(mar= c(5.1, 4.1, 4.1, 2.1) )

```


```{r,genenames}

library("fgsea")

#gs <- gmtPathways("c5.go.v2024.1.Hs.symbols.gmt")
#gs <- gs[grep("^GOBP",names(gs))] #GO BP only
gs <- gmtPathways("c5.hpo.v2024.1.Hs.symbols.gmt")
#gs <- gmtPathways("msigdb.v2024.1.Hs.symbols.gmt")

#download.file("https://ziemann-lab.net/public/tmp/ensg.tsv",destfile="ensg.tsv")
ensg2hgnc <- read.table("ensg.tsv",header=TRUE)


gmt_import <- function(gmtfile) {
    genesetLines <- strsplit(readLines(gmtfile), "\t")
    genesets <- lapply(genesetLines, utils::tail, -2)
    names(genesets) <- unlist(lapply(genesetLines, head, 1))
    attributes(genesets)$originfile <- gmtfile
    genesets <- genesets[lapply(genesets,length)>0]
    if( any(duplicated(names(genesets))) ) {
        warning("Duplicated gene sets names detected")
    }
    genesets
}

gs <- gmt_import("c5.hpo.v2024.1.Hs.symbols.gmt")

#ensg <- rownames(cordata)
#library(biomaRt)
#ensembl <- useEnsembl(biomart = "genes", dataset = "hsapiens_gene_ensembl")
#searchAttributes(mart = ensembl, pattern = "hgnc")
#head(searchAttributes(mart = ensembl, pattern = "stable"))
#head(ensg)
#length(ensg)
#bm <- getBM(attributes = c('ensembl_gene_id', 'hgnc_symbol'),
#      values = ensg, mart = ensembl)
#write.table(bm,"ensg.tsv",sep="\t",row.names=FALSE)


#cordata <- read.table("cordata.tsv.a")

myora <- function(cordata,ensg2hgnc,gs) {
  cm <- merge(ensg2hgnc,cordata,by.x="ensembl_gene_id",by.y=0)
  ores <- mclapply(1:ncol(cordata), function(i) {
    geneid <- colnames(cordata)[i]
    geneid <- paste((ensg2hgnc[grep(geneid,ensg2hgnc$ensembl_gene_id),]),collapse=" ")
    g <- cordata[,i]
    names(g) <- cm[,2]
    g <- g[which(!is.na(g))]
    g <- g[!is.na(names(g))]
    g <- g[which(names(g)!="")]
    g[g == 1] <- 0
    g <- g[order(-g)]
    fg <- unique(names(head(g[which(g>0.4)],100)))
    #fg <- unique(names(g[which(g>0.4)]))
    if ( length(fg) > 50 ) {
      bg <- unique(names(g))
      ores <- fora(pathways=gs,genes=fg,universe=bg,minSize=2,maxSize=100)
      ores$FES <- (ores$overlap / ores$size ) / ( length(fg) / length(bg) )
      ores <- ores[order(-ores$FES),]
      ores <- subset(ores,padj<0.05 & FES > 2 )
    } else {
      ores <- 0
    }
    return(ores)
  },mc.cores=1)
  onames <- lapply(1:ncol(cordata), function(i) {
    geneid <- colnames(cordata)[i]
    geneid <- paste((ensg2hgnc[grep(geneid,ensg2hgnc$ensembl_gene_id),]),collapse=" ")
    geneid
  })
  names(ores) <- onames
  ores <- ores[which(unlist(lapply(ores,object.size) ) > 2296)]
  dfres <- lapply(1:length(ores), function(i) { data.frame( names(ores)[i], ores[[i]] ) } )
  dfres <- do.call(rbind,dfres)
  dfres$overlapGenes <- unlist(lapply(dfres$overlapGenes, function(x) {
    paste(x,collapse="," )
  } ) )
  return(dfres)
}


cordatalist <- list.files(".",pattern="^cordata.tsv.")

ores <- lapply(cordatalist, function(x) {
  outfile=gsub("cordata","oradata",x)
  cordata <- read.table(x,header=TRUE)
  rn <- row.names(cordata)
  cordata <- apply(cordata,2,as.numeric)
  rownames(cordata) <- rn
  oradat <- myora(cordata=cordata,ensg2hgnc,gs)
  write.table(x=oradat,file=outfile,sep="\t",quote=FALSE,row.names=FALSE)
})

```

```{bash,collateresult}

head -1 oradata.tsv.a > impute_res.tsv
grep -v 'names.ores..i.' oradata.tsv.* | cut -d ':' -f2- \
| sed 's/ /_/' | sort -k8gr | sort -k2 >> impute_res.tsv

```



```{r,fcs}

myfcs <- function(cordata,ensg2hgnc,gs) {
  cm <- merge(ensg2hgnc,cordata,by.x="ensembl_gene_id",by.y=0)
  fres <- mclapply(1:ncol(cordata), function(i) {
    geneid <- colnames(cordata)[i]
    geneid <- paste((ensg2hgnc[grep(geneid,ensg2hgnc$ensembl_gene_id),]),collapse=" ")
    g <- cordata[,i]
    names(g) <- cm[,2]
    g <- g[which(!is.na(g))]
    g <- g[!is.na(names(g))]
    g <- g[which(names(g)!="")]
    g[g == 1] <- 0
    gn <- names(g)
    g <- as.vector(scale(g))
    names(g) <- gn
    fres <- fgsea(pathways=gs,stats=g,minSize=3)
    fres <- fres[order(-abs(fres$ES)),]
    fres <- subset(fres,padj<0.01 & ES > 0.4 )
    return(fres)
  },mc.cores=1)
  fnames <- lapply(1:ncol(cordata), function(i) {
    geneid <- colnames(cordata)[i]
    geneid <- paste((ensg2hgnc[grep(geneid,ensg2hgnc$ensembl_gene_id),]),collapse=" ")
    geneid
  })
  names(fres) <- fnames
  fres <- fres[unlist(lapply(fres,function(x) { nrow(x) > 0 } ) )]
  dfres <- lapply(1:length(fres), function(i) {
    data.frame(names(fres)[i], fres[[i]] )
  } )
  dfres <- do.call(rbind,dfres)
  dfres$overlapGenes <- unlist(lapply(dfres$overlapGenes, function(x) {
    paste(x,collapse="," )
  } ) )
  return(dfres)
}

cordatalist <- list.files(".",pattern="^cordata.tsv.")

fres <- lapply(cordatalist[1], function(x) {
  outfile=gsub("cordata","fcsdata",x)
  #x="tmpcordata.tsv.a"
  cordata <- read.table(x,header=TRUE)
  cordata <- cordata[,1:100]
  rn <- row.names(cordata)
  cordata <- apply(cordata,2,as.numeric)
  rownames(cordata) <- rn
  fcsdat <- myfcs(cordata=cordata,ensg2hgnc,gs)
  write.table(x=fcsdat,file=outfile,sep="\t",quote=FALSE,row.names=FALSE)
})

```


```{r,fcs2}

anofcs <- function(g,gs,minSize=5,threads=1) {
  anores <- mclapply(1:length(gs),function(j) {
    g <- g[order(-g)]
    myset <- gs[[j]]
    mysetname <- names(gs)[j]
    inset <- g[which(names(g) %in% myset)]
    insetlen <- length(inset)
    if ( insetlen >= minSize ) {
      notinset <- g[which(! names(g) %in% myset)]
      tres <- t.test(inset,notinset)
      la <- names(inset[which(inset > mean(inset))])
      if (length(la)>20) { la <- la[1:20] }
      la <- paste(la,collapse=",")
      result <- c("setName"=mysetname, "no. members"=insetlen, "mean"=mean(inset),"pval"=tres$p.val , "leadingEdge"=la)
      return(result)
    }
  } , mc.cores=threads )
  anores2 <- do.call(rbind,anores)
  la <- anores2[,"leadingEdge"]
  rownames(anores2) <- anores2[,1]
  anores2 <- t(apply(anores2[,2:4],1,as.numeric))
  colnames(anores2) <- c("no. members","mean","pval")
  anores2 <- as.data.frame(anores2)
  anores2$fdr <- p.adjust(anores2$pval,method="fdr")
  anores2$leadingEdge <- la
  return(anores2)
}

myfcs2 <- function(cordata,ensg2hgnc,gs,threads=1) {
  cm <- merge(ensg2hgnc,cordata,by.x="ensembl_gene_id",by.y=0)
  fres <- mclapply(1:ncol(cordata), function(i) {
    geneid <- colnames(cordata)[i]
    geneid <- paste((ensg2hgnc[grep(geneid,ensg2hgnc$ensembl_gene_id),]),collapse=" ")
    g <- cordata[,i]
    names(g) <- cm[,2]
    g <- g[which(!is.na(g))]
    g <- g[!is.na(names(g))]
    g <- g[which(names(g)!="")]
    g[g == 1] <- 0
    gn <- names(g)
    g <- as.vector(scale(g))
    names(g) <- gn
    if ( length (g) > 5000 ) {
      ares <- anofcs(g,gs,minSize=5,threads=1)
      ares <- subset(ares,fdr<0.01 & mean > 0.3 )
      ares <- ares[order(-ares$mean),]
      return(ares)
    } else {
      return(0)
    }
  },mc.cores=threads)
  fnames <- lapply(1:ncol(cordata), function(i) {
    geneid <- colnames(cordata)[i]
    geneid <- paste((ensg2hgnc[grep(geneid,ensg2hgnc$ensembl_gene_id),]),collapse=" ")
    geneid
  })
  names(fres) <- fnames
  fres <- fres[which(lapply(fres, function(x) { object.size(x) }  ) > 1232)]
  if ( length(fres) > 0 ) {
    dfres <- lapply(1:length(fres), function(i) {
      data.frame( names(fres)[i], fres[[i]] )
    } )
    dfres <- do.call(rbind,dfres)
    return(dfres)
  } else {
    message("No associations found")
  }
}
# 8 core is faster than 16,14,12

tic()
fres <- lapply(cordatalist, function(x) {
  outfile=gsub("cordata","fcsdata",x)
  cordata <- read.table(x,header=TRUE)
  rn <- row.names(cordata)
  cordata <- apply(cordata,2,as.numeric)
  rownames(cordata) <- rn
  fcsdat <- myfcs2(cordata=cordata,ensg2hgnc,gs=gs,threads=4)
  write.table(x=fcsdat,file=outfile,sep="\t",quote=FALSE,row.names=TRUE)
})
toc()

```

Collate.

```{bash,fcscollate1}

head -1 fcsdata.tsv.a | sed 's/^/set\t/' > imputefcs_res.tsv
grep -v 'names.fres..i.' fcsdata.tsv.* | cut -d ':' -f2- \
| sed 's/ /_/' | sort -k4gr | sort -k1 >> imputefcs_res.tsv


cut -f1 imputefcs_res.tsv | sed 's/[0-9]$//g' |sed 's/[0-9]$//g' | sed 's/[0-9]$//g' \
| paste - imputefcs_res.tsv | cut -f1,3- > imputefcs_res.tsv2
mv imputefcs_res.tsv2 imputefcs_res.tsv

```


```{r,data}

linc <- ores[grep("LINC",names(ores))]
linc <- linc[which(sapply(linc,nrow)>1)]
lincnames <- names(which(unlist(lapply(linc,nrow))>0))
linc <- linc[which(names(linc) %in% lincnames)]

names(linc)[1]
head(linc[[1]],10)

names(linc)[2]
head(linc[[2]],10)

names(linc)[3]
head(linc[[3]],10)

names(linc)[4]
head(linc[[4]],10)

names(linc)[5]
head(linc[[5]],10)



ores <- ores[which(unlist(lapply(ores,object.size) ) > 2296)]

ores[grep("ENSG00000284642",names(ores))]

unlist(lapply(ores,nrow))

```

ENSG00000081721 DUSP12 glucokinase regulation confirmed.

ENSG00000117091 CD48 GOBP_B_CELL_RECEPTOR confirmed

ENSG00000118733 OLFM3 GOBP_POSITIVE_REGULATION_OF_NEUROTRANSMITTER_SECRETION


